\LARGE \bf Machine Translation for Tibetan Buddhist Texts

### 2021-04-16 paper/main.tex
Update main.tex
Last commit. Added better conclusion

Our project aims to increase the accessibility of Tibetan religious texts by training a machine translation model specifically to translate religious Tibetan texts to English. Tibetan religions, in particular Buddhism, have a long tradition of translation of texts. While this work was not created within the context of theses traditions, it was certainly inspired by them and we hope it inspires more such translations.

We have trained our models in PyTorch and have leveraged the BART architecture in the HuggingFrance transformers library to make the training and architecture more straightforward. By creating these models, we hope to motivate scholars to study not only these texts but Tibetan as a whole. Having a readily available set of translated discourses could create an on-ramp for further language uses because increased access could lead to increased interest, especially in the English-speaking community. 
Currently, while there has been some interest in machine translation for Tibetan-Chinese and vice versa, the same interest has not been reflected in Tibetan-English machine translation. We attempted to find open-source Tibetan NLP material, but we were unable to find any projects attempting to translate Tibetan-English nor could we find papers specifically discussing Tibetan-English machine translation.

While our model tackles this difficult task in creating a translator for Tibetan-English, it is by no means a perfect or even accurate model. In the final iteration of our model, we were only able to achieve a 23.01\% BLEU score, which means that our model has low-quality output. This is caused by several factors, from the difficulty to preprocess Tibetan sentences to the lack of Tibetan-English data freely available online and our overall knowledge of the Tibetan language. While it may seem that we were unable to fully accomplish our lofty goal of creating a Tibetan-English translator it’s important to note that we are treading in nearly uncharted territory. We hope that our endeavor may help future projects in this field and provide insight into the current state of Tibetan-English machine translation.

@@

As a whole, not much work has been done in the area of Machine Translation or even Natural Language Processing as a whole for the Tibetan Language. There are a few papers exist which concern morphological segmentation and POS tagging for the Tibetan language \cite{c3} \cite{c4}. In recent years there has considerable interest in Chinese to Tibetan Machine translation \cite{c1} \cite{c2}. For example, “A Chinese to Tibetan Machine Translation System with Multiple Translating Strategies” discusses an approach of machine translation from Tibetan to Chinese. The paper makes 3 different translations models, Ruled Based MT, Example Based MT, and Statistical Based MT,  and compares the results. Noticeably missing from this is a Neural Network MT. However, Neural Network MT wasn’t prevalent at that time of its authorship. Nonetheless, the paper concludes, given this data (after processing), that the Statistical Based MT (implemented with conditional random fields) is indeed the best performing option. This is perhaps to be expected. However, as is alluded to above it is unknown whether a Neural Network MT would give a better performance here.

@@

No such papers exist for Tibetan to Enlgish Machine transltions. In fact, to our knowledge this is the first such Tibetan to English Machine translation system developed.  Nonetheless, the findings from the Chinese to Tibetan machine translation have given us a starting point for our investitagion.

@@

To train our model, we acquired publicly available English to Tibetan parallel corpora. The non-profit 84000 provided 80 English to Tibetan parallel texts on their website \cite{c8}. All 80 of these texts were expert translations of Indo-Tibetan Buddhist texts originally written in Tibetan. For our model we used all 80 of these provided texts.

Before using 84000's corpora however, we pre-processed the provided data to meet our model's need. The provided files from 84000 were in a .tmx file format, a very standard parallel corpus file format \cite{c9}. 
For our approach, we scraped all 80 .tmx files for both its English and Tibetan contents and placed them line by line into two large .txt files. One .txt file contained the English lines, and the other .txt file contained the Tibetan translation. We made careful assurance that the lines remained in exactly one to one correspondence with their translations. That is, a newline character was used to distinguish two separate lines of translation, but the same line number in each file corresponds to the exact translation provided by 84000. Therefore, the total line numbers in the two .txt files were necessarily the same since one line in one file corresponding to exactly the same line of translated text in the other file. Two such .txt file were made for each of the 80 .tmx files, a Tibetan and an English .txt file for each. After all 160 .txt files were generated, two single large .txt file were generated by concatenating all the Tibetan .txt files and all the English .txt files. These last two .txt files are what was fed into the model.

It was not strictly necessary to separate the creation of the large .txt files into the smaller components. However, giving the limited data we had this method proved useful to see how much individual texts contributed to the model.

\subsection{Model Design}
We used a 6 layer Transformer model that used a multi headed attention mechanism and a switchout transform. Each layer of the transformer model can be divided into an encoder and decoder layer. The encoding layer consists first of a self attention layer, an attention aspect which focuses on the encoded sentence, and then outputs to a feed forward layer. The multi part of the multi headed attention means that each head will encode the information in parallel and then the results will be concatenated and passed to the decoder. Figure one demonstrates this process.
\includegraphics[scale=.4]{Transformer.png}
\caption{\\Figure 1: Overview of a Transformer model \cite{c11}}
\\
\par
To prevent overfitting and to augment the data, we added in a switchout transform which will randomly switch out words in the source and target sentence. This also has the added advantage of augmenting the data even further which is especially important for our problem as religious Tibetan has a relatively small corpus, around 30 thousand words. We also attempted to use masking, however we found it to be ineffective. Masking hides tokens from the model by replacing the tokens with a <mask> token. Tibetan tokens are not separated in the same way that Latin languages separate their tokens. For example,
a line in our test corpus translated to “Having the altruistic intention of excellent bodhicitta.” However, the Tibetan word is only one token. Therefore, the mask would not help the model learn at all, as the entire sentence would be marked as <mask> and there would not be enough context for our model to discover the meaning. As a result, we did not use any form of masking in our model. Switchout does not create the same problem as their is always a token to take its place. \par
After implementing our transformer model with switchout we tuned hyper parameters to maximize our Transformer model. We manually tuned the following hyperparameters: dropout, label smoothing, rnn size, the number of transformer layers and the number of heads of the attention model.  \par
We used four different metrics to evaluate our results: accuracy, predicted score, perplexity and bilingual evaluation understudy (BLEU). The validation accuracy tells us how well we are learning our training data, the predicted score tells us the average log likelihood for each word, the perplexity tells us how well the model predicts the data and the bleu score evaluates the quality of the translation. Together, these metrics provide us with a decent picture of how well our model is performed.

@@

\par
Our six layer transformer model with switch out was unsuccessful at translating religious Tibetan to English. The translated output does not match closely enough in either pure accuracy or in the meaning of the passage to be useful in translating religious Tibetan texts to English. \par
While superficially, the perplexity appears to demonstrate a strong model, further investigation demonstrates that the low perplexity is a sign of inaccuracy. Perplexity measures the number of possible words that a model could predict for any given sentence. A higher perplexity indicates higher uncertainty. However, the low BLEU score, which measures the quality of the machine translation, has already demonstrated that the accuracy is low. This combined with the perplexity shows the models overconfidence in predicting certain phrases. For example, one Tibetan phrases expert translation was “So the bodhisattva mahāsattva Maitreya asked the Blessed One”, but the top five of the predicted translations all are of the form “The Blessed One then said to the [adj] [proper noun]”. Even though this sentence is wrong, the perplexity is low because the model is confident in the first part of the sentence. This phenomenon is repeated throughout the results. Keyphrases from the Tibetan sentence are interpreted wrongly in the same way, and then the model learns to be confident about its predictions which lowers the perplexity. \par
The results demonstrate the flaw in the Transformer model using the current parallel Tibetan to English corpuses. Each token in the Tibetan sentence corresponds to an entire phrase in English rather than a separate word. This complicates the translation, and the model learns parts of the phrases quite well. The validation accuracy is relatively high because the transformer model is good at learning parts of these phrases but it is not good at learning entire sentences or stringing together sentences in context which contributes to the low BLEU score. Overall, the model is capable of learning specific phrases well, but fails at the broad task of creating coherent translations from religious Tibetan texts to English. \par
We are scheduled to discuss these results with our native Tibetan speaker, Palden Gyal, but unfortunately did get to meet in time for this publication.

@@

As mentioned in the beginning, developing a machine learning model for Tibetan Buddhist texts is a challenging case study primarily due to the extremely low resources available. We overcame many obstacles while training our model and even employed the usefulness of native speakers along with publicly available resources, but our 6 layer transformer model with switch out was unsuccessful at translating religious Tibetan to English. The translated output does not match closely enough in either pure accuracy or in the meaning of the passage to be useful in translating religious Tibetan texts to English, as expanded upon in the results section. Additionally, we explored the ethical considerations of translating such texts and deeply reflected on those, amongst our procedure and findings.

@@

The authors would like to thank Palden Gyal for volunteering to read our translated texts. We would also like to thank Smara Muresan, Isabelle Zuagg, and Jonathan Reeve for mentoring this research. Lastly, we would like to thank the 84000 project for providing open source parallel Tibetan corpora.

>>>>

Our project aims to increase the accessibility of Tibetan religious texts by training a machine translation model specifically to translate religious Tibetan texts to English. Tibetan religions, in particular Buddhism, have a long tradition of translation of texts. While this work was not created within the context of theses traditions, it was certainly inspired by them and we hope it will increase accessibility to Tibetan religious texts and inspire others do undergo similar projects.

We have created our Tibetan to English machine translation models in PyTorch and have leveraged the BART architecture in the HuggingFrance transformers library to make the training and architecture more straightforward. By creating these models, we hope to motivate scholars to study not only these texts but Tibetan as a whole. Having a readily available set of translated discourses could create an on-ramp for further language uses because increased access could lead to increased interest, especially in the English-speaking community. 
Currently, while there has been some interest in machine translation for Tibetan-Chinese and vice versa, the same interest has not been reflected in Tibetan-English machine translation. We attempted to find open-source Tibetan NLP material, but we were unable to find any projects attempting to translate Tibetan-English nor could we find papers specifically discussing Tibetan-English machine translation.

While our model tackles this difficult task in creating a translator for Tibetan-English, it is by no means a perfect or even accurate model. In the final iteration of our model, we were only able to achieve a 23.01\% BLEU score, which means that our model has low-quality output. This is caused by several factors, from the difficulty to preprocess Tibetan sentences to the lack of Tibetan-English data freely available online and our overall knowledge of the Tibetan language. While it may seem that we were unable to fully accomplish our lofty goal of creating a Tibetan-English translator it is important to note that we are treading in nearly uncharted territory. We hope that our endeavor may help future projects in this field and provide insight into the current state of Tibetan-English machine translation.

@@

As a whole, not much work has been done in the area of machine translation (MT) or even Natural Language Processing (NLP) as a whole for the Tibetan language. There are a few papers exist which concern morphological segmentation and parts of speech (POS) tagging for the Tibetan language \cite{c3} \cite{c4}. In recent years there has considerable interest in Chinese to Tibetan machine translation \cite{c1} \cite{c2}. For example, “A Chinese to Tibetan Machine Translation System with Multiple Translating Strategies” discusses an approach of machine translation from Tibetan to Chinese. The paper makes 3 different translations models, Ruled Based MT, Example Based MT, and Statistical Based MT,  and compares the results. Noticeably missing from this is a Neural Network MT. However, Neural Network MT wasn’t prevalent at that time of its authorship. Nonetheless, the paper concludes, given this data (after processing), that the Statistical Based MT,implemented with conditional random fields, is indeed the best performing option. This is perhaps to be expected. However, as is alluded to above it is unknown whether a Neural Network MT would give a better performance here.

@@

Unfortunately, no such papers exist for Tibetan to English machine translation. In fact, to our knowledge this is the first such Tibetan to English machine translation system developed.  Nonetheless, the findings from the Chinese to Tibetan machine translation have given us a starting point for our investigation.

@@

To train our model, we acquired publicly available English to Tibetan parallel corpora. The non-profit 84000 provided 80 English to Tibetan parallel texts on their website \cite{c8}. All 80 of these texts were expert translations of Indo-Tibetan Buddhist texts originally written in Tibetan. For our model we used all 80 of these provided texts. 84000 is still actively creating more translated texts, so our model was only trained on to their most recent release which from the time of this papers' publication was March 2021.

Before using 84000's corpora however, we pre-processed the provided data to meet our model's need. The provided files from 84000 were in a .tmx file format, a very standard parallel corpus file format \cite{c9}. 
For our approach, we scraped all 80 .tmx files for both its English and Tibetan contents and placed them line by line into two large .txt files. One .txt file contained the English lines, and the other .txt file contained the Tibetan translation. We made careful assurance that the lines remained in exactly one to one correspondence with their translations. That is, a newline character was used to distinguish two separate lines of translation, but the same line number in each file corresponds to the exact translation provided by 84000. Therefore, the total line numbers in the two .txt files were necessarily the same since one line in one file corresponds to exactly the same line of translated text in the other file. Two such .txt file were made for each of the 80 .tmx files, a Tibetan and an English .txt file for each. After all 160 .txt files were generated, two large .txt file were generated by concatenating all the Tibetan .txt files and all the English .txt files. These last two large .txt files are what was fed into the model. The first few lines of our two large .txt files are shown below in figure 1.

Note in the the figure, the left hand side represents the source language (Tibetan) expressed in the Tibetan Unicode standard. On the right is the expert English translation provided by 84000. There is a one to one mapping between lines. This is, the Tibetan text on line one exactly corresponds to the English on line one. Of course, these are two separate files, and this picture simply reflects the view of the first few lines as would be seen if opened in a simple text editor.

Note that it was not strictly necessary to separate the creation of the large .txt files into the smaller components. However, giving the limited data we had this method proved useful to see how much individual texts contributed to the model.

\begin{figure*}
  \includegraphics[width=\textwidth, height=6cm]{Screen Shot 2021-04-16 at 6.40.00 PM.png}
  \caption{The first few lines of our fully concatenated parallel corpus}
\end{figure*}

\subsection{Model Design}
We used a 6 layer Transformer model that used a multi headed attention mechanism and a switchout transform. Each layer of the transformer model can be divided into an encoder and decoder layer. The encoding layer consists first of a self attention layer, an attention aspect which focuses on the encoded sentence, and then outputs to a feed forward layer. The multi part of the multi headed attention means that each head will encode the information in parallel and then the results will be concatenated and passed to the decoder. Figure one demonstrates this process.
\includegraphics[scale=.4]{Transformer.png}
\caption{\\Fig 2. Overview of a Transformer model \cite{c11}}
\\
\par
To prevent overfitting and to augment the data, we added in a switchout transform which will randomly switch out words in the source and target sentence. This also has the added advantage of augmenting the data even further which is especially important for our problem as religious Tibetan has a relatively small corpus, around 30 thousand words. We also attempted to use masking, however we found it to be ineffective. Masking hides tokens from the model by replacing the tokens with a <mask> token. Tibetan tokens are not separated in the same way that Latin languages separate their tokens. For example,
a line in our test corpus translated to “Having the altruistic intention of excellent bodhicitta.” However, the Tibetan word is only one token. Therefore, the mask would not help the model learn at all, as the entire sentence would be marked as <mask> and there would not be enough context for our model to discover the meaning. As a result, we did not use any form of masking in our model. Switchout does not create the same problem as their is always a token to take its place. \par
After implementing our transformer model with switchout we tuned hyper parameters to maximize our Transformer model. We manually tuned the following hyperparameters: dropout, label smoothing, rnn size, the number of transformer layers and the number of heads of the attention model.  \par
We used four different metrics to evaluate our results: accuracy, predicted score, perplexity and bilingual evaluation understudy (BLEU). The validation accuracy tells us how well we are learning our training data, the predicted score tells us the average log likelihood for each word, the perplexity tells us how well the model predicts the data and the bleu score evaluates the quality of the translation. Together, these metrics provide us with a decent picture of how well our model is performed.

@@

\par
Our six layer transformer model with switch out was unsuccessful at translating religious Tibetan to English. The translated output does not match closely enough in either pure accuracy or in the meaning of the passage to be useful in translating religious Tibetan texts to English. \par
While superficially, the perplexity appears to demonstrate a strong model, further investigation demonstrates that the low perplexity is a sign of inaccuracy. Perplexity measures the number of possible words that a model could predict for any given sentence. A higher perplexity indicates higher uncertainty. However, the low BLEU score, which measures the quality of the machine translation, has already demonstrated that the accuracy is low. This combined with the perplexity shows the models overconfidence in predicting certain phrases. For example, one Tibetan phrases expert translation was “So the bodhisattva mahāsattva Maitreya asked the Blessed One”, but the top five of the predicted translations all are of the form “The Blessed One then said to the [adj] [proper noun]”. Even though this sentence is wrong, the perplexity is low because the model is confident in the first part of the sentence. This phenomenon is repeated throughout the results. Some examples of a few of these cases are shown in Figure 3.

\newline \,\,
\includegraphics[scale=.36]{Screen Shot 2021-04-16 at 6.54.34 PM.png}
\caption{\\Fig 3. Example of problematic predictions }
\newline \,\,

Keyphrases from the Tibetan sentence are interpreted wrongly in the same way, and then the model learns to be confident about its predictions which lowers the perplexity. \par
The results demonstrate the flaw in the Transformer model using the current parallel Tibetan to English corpuses. Each token in the Tibetan sentence corresponds to an entire phrase in English rather than a separate word. This complicates the translation, and the model learns parts of the phrases quite well. The validation accuracy is relatively high because the transformer model is good at learning parts of these phrases but it is not good at learning entire sentences or stringing together sentences in context which contributes to the low BLEU score. Overall, the model is capable of learning specific phrases well, but fails at the broad task of creating coherent translations from religious Tibetan texts to English. \par
We are scheduled to discuss these results and possible workarounds to our problems with our native Tibetan speaker, Palden Gyal, but unfortunately did get to meet in time for this publication.

@@

As mentioned in our introduction, developing a machine learning model for Tibetan Buddhist texts is a challenging task primarily due to the extremely low resources available. We overcame many obstacles while training our model  while following a well defined ethical approach and using only  publicly available resources. However, our 6 layer transformer model with switch out was unsuccessful at translating religious Tibetan to English. The translated output does not match closely enough in either pure accuracy or in the meaning of the passage to be useful in translating religious Tibetan texts to English, as expanded upon in the results section. \par
Additionally, we explored the ethical considerations of translating such texts and deeply reflected on those, amongst our procedure and findings. We faced an ethical dilemma in this part of our research project as to whether certain religious texts should or should not be translated given their sacredness. Another hurdle we overcame was employing switchout over masking as masking tends to hide tokens. We also realized that initially the result for perplexity seemed to demonstrate a strong model but further investigation brought forth that lower perplexity was actually a sign of accuracy.  To be able to do a real world sense check and valuable input, we have scheduled a meeting with a native Tibetan speaker, through which we hope to be able to further enhance our model and increase accuracy of the same.  \par
To conclude, while our work is not successful, using transformer models might show promise if new approaches can find a way to add more training data or stop repeated tokens along with insights that may come from completing primary diligence in the form of conversations with native speakers.

@@

The authors would like to thank Palden Gyal for volunteering to read our translated texts. We would also like to thank Smara Muresan, Isabelle Zaugg, and Jonathan Reeve for mentoring this research. We would like to thank Professor Thomas Yarnall and Professor Paul Hackett for providing additional advice. Lastly, we would like to thank the 84000 project for providing open source parallel Tibetan corpora.
